---
---
---

# Lund University BIOS14 Final Exam

## Jacob Daigle \| 2025-01-15

For this exam, we are given a large dataset concerning a field monitoring project studying two species of European riverine damselfly, *Calopteryx splendens*, and *Calopteryx virgo*.

The questions I am interested in attempting to explore using this dataset are as follows:

1.  Forewing patch length and width (and thus area) are supposedly a target of sexual selection. Is this reflected in this dataset? How does an increase in patch area affect the probability an individual is to have copulated?
2.  What is the relationship between body size and wing size, taking into account differences in sexes and species?
3.  How have the different body segment lengths changed over time?

```{r}
rm(list = ls())
library(dplyr)

male_CS <- read.csv("../Data/male_CS.csv")
male_CV <- read.csv("../Data/male_CV.csv")
female_CS <- read.csv("../Data/female_CS.csv")
female_CV <- read.csv("../Data/female_CV.csv")
dat <- bind_rows(male_CS, male_CV, female_CS, female_CV)
dat[["fpa"]] <- dat[["fpl"]] * dat[["fpw"]]

logit = function(x) log(x/(1 - x))
invlogit = function(x) 1/(1 + exp(-x))
```

# Forewing Patch Area and Copulation Status

An increase of one standard deviation of forewing patch area is associated with a *drop* in likelihood to copulate by 0.9%. The coefficient of discrimination calculated to be 0.0022.

```{r}
plot(dat$fpa, dat$cop, main = "Copulation Likelihood vs. Forewing Patch Area",
     xlab = "Forewing Patch Area (mm^2)", ylab = "Copulation Likelihood")

# Logistic regression, binary response variable
cop_by_fpa <- glm(formula = cop ~ fpa, family = binomial(link = "logit"), data = dat)
xvals <- seq(min(dat$fpa, na.rm = T), max(dat$fpa, na.rm = T), 0.1)
coefs <- cop_by_fpa$coefficients
yvals <- coefs[1] + coefs[2]*xvals

lines(xvals, invlogit(yvals))

# Calculates difference between mean and mean + 1SD
print(
    invlogit(
        coefs[1] + coefs[2] * (mean(dat$fpa, na.rm = T) + sd(dat$fpa, na.rm = T))
    ) - invlogit(
        coefs[1] + coefs[2] * (mean(dat$fpa, na.rm = T))
    )
)
# Calculates coefficient of discrimination
print(
    mean(invlogit(yvals)[which(dat$cop == 1)], na.rm = T)
         - mean(invlogit(yvals)[which(dat$cop == 0)], na.rm = T)
    )
```

# Forewing Length and Body Size

The variance in forewing is partitioned as follows:

|                                |                              |
|--------------------------------|------------------------------|
| Parameter                      | Amount of Variance Explained |
| Total body length              | 5.2%                         |
| Species                        | 3.9%                         |
| Sex                            | 45.5%                        |
| Interaction of sex and species | 0.7%                         |
| Residuals (unexplained)        | 44.6%                        |

We expect wing length to relate allometrically to total body length, so $winglength = a\times bodylength^b$. We can take the log of both sides to be able to fit a linear regression to the data, so $\text{log}(winglength) = \text{log}(a) + b\times\text{log}(bodylength)$. An ANCOVA analysis is performed twice to see the difference sex and species each have on the relationship.

We find that the difference between species is not at all like the difference between sexes. The allometric slope is $0.68\pm 0.0035\frac{\text{log mm wing length}}{\text{log mm body length}}$ for females, so a 10% increase in total body length is associated with a 6.8% increase in forewing length, while in males a 10% increase in body length is associated with a 6.1% increase in forewing length - not much different. However in *C. Virgo*, a 10% body length increase is associated with a 4.1% increase in forewing length ($0.40803 \pm 0.08236 \frac{\text{log mm wing length}}{\text{log mm body length}}$), while a 10% body length increase in *C. Splendens* is associated with a 1.2% **decrease** in forewing length.

As seen in the graph, though, there does appear to be three subgroups of *C. Splendens*. I'm not sure exactly what the subgroups are, but I decided to redo the ANOVA for this species assuming three differently sized subgroups, which I've labelled "small", "medium", and "large", though of course all are very close in size.

```{r}
# By Sex ----------------------------------
# Log-log is taken so relationship is linear
fwl_by_tbl_sex <- lm(log10(fwl) ~ log10(tbl) * as.factor(sex), data = dat)
coefs <- summary(fwl_by_tbl_sex)$coef

# X values used to graph regression lines within data later
xvals1 <- seq(
    min(log10(dat$tbl[dat$sex == "Female"]), na.rm = T),
    max(log10(dat$tbl[dat$sex == "Female"]), na.rm = T),
    0.1)
xvals2 <- seq(
    min(log10(dat$tbl[dat$sex == "Male"]), na.rm = T),
    max(log10(dat$tbl[dat$sex == "Male"]), na.rm = T),
    0.1)

plot(log10(dat$tbl), log10(dat$fwl),
     main = "Wing Length vs. Total Body Length",
     xlab = "Total Body Length (mm)", ylab = "Wing Length (mm)",
     pch = 16, cex = 1, axes = F,
     col = c(rgb(1, 0, 0, 0.1), rgb(0, 0, 1, 0.1))[(dat$sex == "Male") + 1])
# Axes are drawn manually to get 10^(log value) instead of (log value)
axis(1, at = seq(1.55, 1.75, 0.05),
     labels = parse(text = paste0("10^", seq(1.55, 1.75, 0.05))))
axis(2, at = seq(1.3, 1.6, 0.05), las = 1,
     labels = parse(text = paste0("10^", seq(1.3, 1.6, 0.05))))

# Regression lines (linear)
lines(xvals1, coefs[1, 1] + coefs[2, 1] * xvals1,
      col = "red3", lwd = 3)
lines(xvals2, coefs[1, 1] + coefs[3, 1] + (coefs[2, 1] + coefs[4, 1]) * xvals2,
      col = "blue3", lwd = 3)

legend("bottomright",
       c("Female", "Male"),
       col = c("red3", "blue3"),
       lwd = 3)

# By Species -------------------------------

fwl_by_tbl <- lm(log10(fwl) ~ log10(tbl) * as.factor(sp), data = dat)
coefs <- summary(fwl_by_tbl)$coef

xvals1 <- seq(
    min(log10(dat$tbl[dat$sp == "CS"]), na.rm = T),
    max(log10(dat$tbl[dat$sp == "CS"]), na.rm = T),
    0.1)
xvals2 <- seq(
    min(log10(dat$tbl[dat$sp == "CV"]), na.rm = T),
    max(log10(dat$tbl[dat$sp == "CV"]), na.rm = T),
    0.1)

plot(log10(dat$tbl), log10(dat$fwl),
     main = "Wing Length vs. Body Length",
     xlab = "Total Body Length (mm)", ylab = "Wing Length (mm)",
     pch = 16, cex = 1, axes = F,
     col = c(rgb(255, 150, 0, 20, maxColorValue = 255),
             rgb(0, 255, 0, 20, maxColorValue = 255)
             )[(dat$sp == "CV") + 1])
axis(1, at = seq(1.55, 1.75, 0.05),
     labels = parse(text = paste0("10^", seq(1.55, 1.75, 0.05))))
axis(2, at = seq(1.3, 1.6, 0.05), las = 1,
     labels = parse(text = paste0("10^", seq(1.3, 1.6, 0.05))))

lines(xvals1, coefs[1, 1] + coefs[2, 1] * xvals1,
      col = "orange3", lwd = 3)
lines(xvals2, coefs[1, 1] + coefs[3, 1] + (coefs[2, 1] + coefs[4, 1]) * xvals2,
      col = "green3", lwd = 3)

legend("bottomright",
       c("C. Splendens", "C. Virgo"),
       col = c("orange3", "green3"),
       lwd = 3)

summary(fwl_by_tbl_sex)
summary(fwl_by_tbl)
anova(fwl_by_tbl_sex)
anova(fwl_by_tbl)
```

If we separate *C. Splendens* into three subgroups, we find that for a 10% increase in body length, there is a 3.9%, 2.3%, and 1.9% increase in body length for "small", "medium", and "large" flies respectively. The body length, size category, interaction of body length and size category, and residuals account for 0.3%, 86.0%, 0.1%, and 13.6% of the variance respectively.

```{r}
# Subgrouping -------------------------------
subdat <- dat[which(dat$sp == "CS"),]  # Data only of C. Splendens
subdat$size <- factor(cut(
    log10(subdat$fwl),
    c(-Inf, 1.45, 1.5, Inf),
    labels = c("small", "medium", "large")
))

fwl_by_tbl_sub <- lm(log10(fwl) ~ log10(tbl) * size, data = subdat)
coefs <- summary(fwl_by_tbl_sub)$coef

# X values for drawing regression lines within data
xvals1 <- seq(
    min(log10(subdat$tbl[subdat$size == "small"]), na.rm = T),
    max(log10(subdat$tbl[subdat$size == "small"]), na.rm = T),
    0.01)

xvals2 <- seq(
    min(log10(subdat$tbl[subdat$size == "medium"]), na.rm = T),
    max(log10(subdat$tbl[subdat$size == "medium"]), na.rm = T),
    0.01)

xvals3 <- seq(
    min(log10(subdat$tbl[subdat$size == "large"]), na.rm = T),
    max(log10(subdat$tbl[subdat$size == "large"]), na.rm = T),
    0.01)

plot(log10(subdat$tbl), log10(subdat$fwl),
     main = "Wing Length vs. Body Length in C. Splendens",
     xlab = "Total Body Length (mm)", ylab = "Wing Length (mm)",
     pch = 16, cex = 1, axes = F,
     col = c(
         rgb(1, 0, 0, 0.05),
         rgb(0, 1, 0, 0.05),
         rgb(0, 0, 1, 0.05)
     )[subdat$size]
)
box()  # Draws box around the plot

# Axes with 10^(log val) instead of just (log val)
axis(1, at = seq(1.58, 1.7, 0.02),
     labels = parse(text = paste0("10^", seq(1.58, 1.7, 0.02))))
axis(2, at = seq(1.3, 1.6, 0.05), las = 1,
     labels = parse(text = paste0("10^", seq(1.3, 1.6, 0.05))))

# Regression lines
lines(xvals1,
      coefs[1,1] + coefs[2,1] * xvals1,
      col = "red", lwd = 2)  # Small
lines(xvals2,
      coefs[1,1] + coefs[3,1] + (coefs[2,1] + coefs[5,1]) * xvals2,
      col = "green", lwd = 2)  # Medium
lines(xvals3,
      coefs[1,1] + coefs[4,1] + (coefs[2,1] + coefs[6,1]) * xvals3,
      col = "blue", lwd = 2)  # Large

legend("bottomright",
       c("Small", "Medium", "Large"),
       col = c("red", "green", "blue"),
       lwd = 2)

summary(fwl_by_tbl_sub)
anova(fwl_by_tbl_sub)
```

# Segment Lengths Over Time

The histograms below show a slight decrease in body length over time, $-0.46\frac{mm}{yr}(\pm 0.023)$, but a large increase in thorax length over time, $+1.2\frac{mm}{yr}(\pm 0.010)$. We can also see that the abdomens show a similar decrease over time, $-0.23\frac{mm}{yr}(\pm 0.020)$, which means the heads must be shrinking enormously! The second chart shows the median body lengths over time along with the sum of the median thorax and abdomen lengths.

```{r}
plot(as.factor(dat$year), dat$tbl,
     xlab = "Year", ylab = "Total Body Length (mm)",
     main = "Total Body Length Over Time")
plot(as.factor(dat$year), dat$thorl,
     xlab = "Year", ylab = "Thorax Length (mm)",
     main = "Thorax Length Over Time")
plot(as.factor(dat$year), dat$abl,
     xlab = "Year", ylab = "Abdomen Length (mm)",
     main = "Abdomen Length Over Time")
plot(as.factor(dat$year), dat$tbl - (dat$abl + dat$thorl),
     xlab = "Year", ylab = "Difference Length (mm)",
     main = "'Head' Length Over Time")

summary(lm(tbl ~ as.numeric(year), data = dat))$coef
summary(lm(thorl ~ as.numeric(year), data = dat))$coef
summary(lm(abl ~ as.numeric(year), data = dat))$coef
```

```{r}
# Calculates the median body, abdomen, and thorax lengths for each year
tbl_med <- tapply(dat$tbl, dat$year, median)
abl_med <- tapply(dat$abl, dat$year, median)
thorl_med <- tapply(dat$thorl, dat$year, median)

# Plots the body lengths along with thorax + abdomen
plot(as.numeric(names(tbl_med)), unname(tbl_med), type = "l",
     ylim = c(38, 47), col = "red",
     xlab = "Year", ylab = "Length (mm)",
     main = "Median Total Body Length and Thorax + Abdomen Over Time")
lines(as.numeric(names(abl_med)), unname(abl_med) + unname(thorl_med),
      col = "blue")
legend("bottomright",
       c("Median Total Body Length",
         "Median Thorax + Median Abdomen Length"),
       col = c("red", "blue"), lwd = 2, cex = 0.9)
```
